# Macro-sample Sequencing Statistics

Remember to change M to D, as only microsamples start from M, and these are macro-samples.

## Load required data

Load sample metadata
```{r macro_stats_load_data, message=FALSE, eval=FALSE}
load("data/macro/sample_metadata.Rdata")
load("data/macro/counts.Rdata")
```

## FastP sequencing statistics

Download preprocess_data.zip & place in appropriate directory.

Get the multiqc_fastp.txt: FastP results, including QC. Packed in JSON format.
(Does not contain gc and % unique content for before trimming.)

Load and process FastP sequencing statistics
```{r macro_stats_fastp, warning=FALSE, comments="", message=FALSE, eval=FALSE}
fastp_seq_stats_macro <- read_tsv("data/macro/preprocess_data/multiqc_fastp.txt", 
           col_select = c("Sample", "summary", "duplication", "filtering_result", "adapter_cutting"), 
           show_col_types = FALSE) %>%
  mutate(Sample = str_extract(Sample, "D\\d+"))%>%
  mutate(duplication = as.numeric(gsub(".*:\\s*([0-9\\.]+).*", "\\1", duplication))) %>%
  rename(microsample = Sample) %>%
  mutate(
    summary_clean = gsub("'", "\"", summary),
    total_sequences_before_trim = map_dbl(summary_clean, ~ fromJSON(.)$before_filtering$total_reads),
    total_bases_before_trim = map_dbl(summary_clean, ~ fromJSON(.)$before_filtering$total_bases),
    percent_gc_before_trim = map_dbl(summary_clean, ~ fromJSON(.)$before_filtering$gc_content),
    total_sequences_after_trim = map_dbl(summary_clean, ~ fromJSON(.)$after_filtering$total_reads),
    total_bases_after_trim = map_dbl(summary_clean, ~ fromJSON(.)$after_filtering$total_bases),
    percent_gc_after_trim = map_dbl(summary_clean, ~ fromJSON(.)$after_filtering$gc_content)
  ) %>%
  mutate(
    adaptor_clean = gsub("'", "\"", adapter_cutting),
    adapter_trimmed_reads = map_dbl(adaptor_clean, ~ fromJSON(.)$adapter_trimmed_reads),
    adapter_trimmed_bases = map_dbl(adaptor_clean, ~ fromJSON(.)$adapter_trimmed_bases)
  ) %>%
  mutate(
    filtering_clean = gsub("'", "\"", filtering_result),
    passed_filter_reads = map_dbl(filtering_clean, ~ fromJSON(.)$passed_filter_reads),
    low_quality_reads = map_dbl(filtering_clean, ~ fromJSON(.)$low_quality_reads),
    too_many_N_reads = map_dbl(filtering_clean, ~ fromJSON(.)$too_many_N_reads),
    too_short_reads = map_dbl(filtering_clean, ~ fromJSON(.)$too_short_reads)
  ) %>%
  select(-summary_clean,-summary, -adaptor_clean, -filtering_result, -filtering_clean) %>%
  filter(grepl("^D", microsample)) %>%
  group_by(microsample) %>%
  summarise(
    total_sequences_before_trim = sum(total_sequences_before_trim, na.rm = TRUE),
    total_bases_before_trim = sum(total_bases_before_trim, na.rm = TRUE),
    total_sequences_after_trim = sum(total_sequences_after_trim, na.rm = TRUE),
    total_bases_after_trim = sum(total_bases_after_trim, na.rm = TRUE),
    percent_gc_before_trim = mean(percent_gc_before_trim, na.rm = TRUE)*100,
    percent_gc_after_trim = mean(percent_gc_after_trim, na.rm = TRUE),
    duplication = mean(duplication, na.rm = TRUE)*100,
    adapter_trimmed_reads = sum(adapter_trimmed_reads, na.rm = TRUE),
    adapter_trimmed_bases = sum(adapter_trimmed_bases, na.rm = TRUE),
    passed_filter_reads = sum(passed_filter_reads, na.rm = TRUE),
    low_quality_reads = sum(low_quality_reads, na.rm = TRUE),
    too_many_N_reads = sum(too_many_N_reads, na.rm = TRUE),
    too_short_reads = sum(too_short_reads, na.rm = TRUE)   
    ) %>%
  filter(microsample %in% sample_metadata_macro$sample)
```



## Non-bacterial mapped reads: human, chicken, swine, turkey, histomonas

Get the multiqc_samtools_stats.txt.

This is calculated on the trimmed reads. 
The reads are mapped to these references:
  1. turkey (Turkey_5.1),
  2. histomonas (Hmelea),
  3. chicken (GRCg7b), 
  4. human (GRCh38), and
  5. swine (Sscrofa11.1)) sequentially.

To find the number of reads mapped to each reference you have to either
  - estimate number from the columns: reads_mapped + (reads_mapped - reads_mapped_and_paired) OR
  - subtract the number of reads that entered the analysis (e.g. column 'total_passed' of human - from column 'total_passed' of chicken = reads mapped to human).
  If you try both, these numbers should be equal. 
  
Load and process non-bacterial mapping statistics
```{r macro_stats_non_bacteria_mapping, warning=FALSE, comments="", message=FALSE, eval=FALSE}
stats_non_bacteria_mapping_macro <- read_tsv("data/macro/preprocess_data/multiqc_samtools_stats.txt", show_col_types = FALSE) %>%
  # make new columns
  mutate(reference = case_when(
      # for type of reference based on reference code
      grepl("GRCh38", Sample, ignore.case = TRUE) ~ "human",
      grepl("GRCg7b", Sample, ignore.case = TRUE) ~ "chicken",
      grepl("Sscrofa11.1", Sample, ignore.case = TRUE) ~ "swine",
      grepl("Turkey_5.1", Sample, ignore.case = TRUE) ~ "turkey",
      grepl("Hmelea", Sample, ignore.case = TRUE) ~ "histomonas",
      TRUE ~ NA_character_),
      # for microsample names
      microsample = str_extract(Sample, "D\\d+"),
      # for number of reads as input to each reference mapping
      chicken_total_passed = ifelse(reference == "chicken", raw_total_sequences, NA_real_),
      human_total_passed = ifelse(reference == "human", raw_total_sequences, NA_real_),
      swine_total_passed = ifelse(reference == "swine", raw_total_sequences, NA_real_),
      turkey_total_passed = ifelse(reference == "turkey", raw_total_sequences, NA_real_),
      histomonas_total_passed = ifelse(reference == "histomonas", raw_total_sequences, NA_real_),
      # for number of reads that mapped to each reference 
      chicken_total_mapped = ifelse(reference == "chicken", reads_mapped + (reads_mapped - reads_mapped_and_paired), NA_real_),
      human_total_mapped = ifelse(reference == "human", reads_mapped + (reads_mapped - reads_mapped_and_paired), NA_real_),
      swine_total_mapped = ifelse(reference == "swine", reads_mapped + (reads_mapped - reads_mapped_and_paired), NA_real_),
      turkey_total_mapped = ifelse(reference == "turkey", reads_mapped + (reads_mapped - reads_mapped_and_paired), NA_real_),
      histomonas_total_mapped = ifelse(reference == "histomonas", reads_mapped + (reads_mapped - reads_mapped_and_paired), NA_real_),
      # for unmapped reads after mapping to all three references
      swine_unmapped = ifelse(reference == "swine", swine_total_passed - swine_total_mapped, NA_real_)) %>%
  # select relevant columns
  select(microsample,
        chicken_total_passed, human_total_passed, swine_total_passed, turkey_total_passed, histomonas_total_passed,
        chicken_total_mapped, human_total_mapped, swine_total_mapped, turkey_total_mapped, histomonas_total_mapped,
        swine_unmapped) %>%
  # sum reads for same microsample for each reference
  group_by(microsample) %>%
  summarise(across(
    starts_with("chicken_") | starts_with("human_") | starts_with("swine_")| starts_with("turkey_") | starts_with("histomonas_"), 
    sum, na.rm = TRUE)) %>%
  filter(grepl("^D", microsample)) %>%
  filter(microsample %in% sample_metadata_macro$sample)
```



## Bacterial mapped reads

Download quantify.zip & place in appropriate directory.
Get the multiqc_samtools_stats.txt.

This is calculated on the trimmed reads after filtering for turkey, histomonas, human, chicken, and pig reads.
'Unmapped reads' are trimmed reads that did not map to the turkey, histomonas, human, chicken, and swine references, or to the bacterial MAG catalogue.

Load and process bacterial mapping statistics
```{r macro_stats_bacteria_mapping, warning=FALSE, comments="", message=FALSE, eval=FALSE}
stats_bacteria_mapping_macro <- read_tsv("data/macro/quantify/multiqc_samtools_stats.txt", show_col_types = FALSE) %>%
    # select only the dereplicated to 95% (as the file includes also the 98% and 99%)
  filter(str_detect(Sample, "0\\.95") & str_detect(Sample, "REF0029")) %>%
    # make new columns
  mutate(
    microsample = str_extract(Sample, "D\\d+"),
    bacteria_total_passed = raw_total_sequences,
    bacteria_total_mapped = reads_mapped,
    unmapped = reads_unmapped
  ) %>%
  # sum reads of the same microsample because some samples were sequence multiple times
  group_by(microsample) %>% 
  summarise(
    bacteria_total_passed = sum(bacteria_total_passed, na.rm = TRUE), 
    bacteria_total_mapped = sum(bacteria_total_mapped, na.rm = TRUE),
    unmapped = sum(unmapped, na.rm = TRUE)
  ) %>%
  # select relevant columns
  select(microsample, bacteria_total_passed, bacteria_total_mapped, unmapped) %>%
  # keep microsamples starting with D (i.e., remove 'NA' or 'undetermined')
  filter(grepl("^D", microsample)) 
```



## Alternative way to estimate bacterial read counts (from count table)

Calculate bacterial read counts from count table
```{r macro_stats_bacteria_mapping_alternative, warning=FALSE, comments="", message=FALSE, eval=FALSE}
stats_bacteria_mapping_alternative_macro <- read_counts_macro %>%
  # Move genome names to rownames
  column_to_rownames(var = "genome") %>%
  # Transpose the data
  t() %>% 
  # Convert back to a data frame
  as.data.frame() %>%
  # Move sample names into a column
  rownames_to_column(var = "microsample") %>%
  # sum the counts of all genomes in each sample (i.e., sum of each row)
  mutate(bacteria_total_read_counts = rowSums(select(., -microsample))) %>% 
  # select to show only the total_counts column
  select(microsample, bacteria_total_read_counts) %>% 
  # select only microsamples starting with 'D' (i.e. not 'NA' or 'Undetermined')
  filter(grepl("^D", microsample))
```



## Sanity checks

Check if the two methods of counting bacterial reads agree:

Compare bacterial mapping methods
```{r macro_stats_compare_bacterial_mapping, warning=FALSE, comments="", message=FALSE, eval=FALSE}
comparison_bacteria_mapping_macro <- stats_bacteria_mapping_alternative_macro %>%
  # Join on 'microsample'
  inner_join(stats_bacteria_mapping_macro, by = "microsample") %>% 
  # Compare the two columns
  mutate(same_counts = bacteria_total_read_counts == bacteria_total_mapped) 

# Check if all values are the same
# TRUE if all are the same, FALSE otherwise
all(comparison_bacteria_mapping_macro$same_counts) 

# Filter & show rows where the counts between the two columns differ
comparison_bacteria_mapping_macro %>%
  filter(!same_counts) %>%
  mutate(difference = bacteria_total_read_counts- bacteria_total_mapped) %>%
  select(microsample, bacteria_total_read_counts, bacteria_total_mapped, difference)
```




## Combine relevant sequencing statistics

Combine all sequencing statistics into a single dataframe
```{r macro_stats_combined, warning=FALSE, comments="", message=FALSE, eval=FALSE}
stats_combined_macro <- reduce(list(fastp_seq_stats_macro,
                              stats_non_bacteria_mapping_macro, 
                              stats_bacteria_mapping_macro,
                              stats_bacteria_mapping_alternative_macro), 
                         full_join, by = "microsample") %>%
  filter(grepl("^D", microsample))  %>%
  mutate(removed_sequences_after_trim = total_sequences_before_trim - total_sequences_after_trim)

# estimate some percentages for easier plotting.
final_combined_stats_macro <- stats_combined_macro %>%
  mutate(
    trimmed_reads_percentage = ((total_sequences_before_trim-total_sequences_after_trim)/total_sequences_before_trim)*100,
    human_percentage = (human_total_mapped/total_sequences_after_trim)*100,
    chicken_percentage = (chicken_total_mapped/total_sequences_after_trim)*100,
    swine_percentage = (swine_total_mapped/total_sequences_after_trim)*100,
    turkey_percentage = (turkey_total_mapped/total_sequences_after_trim)*100,
    histomonas_percentage = round((histomonas_total_mapped/total_sequences_after_trim)*100, 7),
    bacteria_percentage = (bacteria_total_mapped/total_sequences_after_trim)*100,
    unmapped_percentage = (unmapped/total_sequences_after_trim)*100,
    bacteria_percentage_enriched = (bacteria_total_mapped/(bacteria_total_mapped+unmapped)*100),
    unmapped_percentage_enriched = (unmapped/(bacteria_total_mapped+unmapped)*100),
    adapter_contamination_pct = (adapter_trimmed_bases / total_bases_before_trim)*100,
    histomonas_over_bacteria = round((histomonas_total_mapped/bacteria_total_mapped),7),
    bacteria_over_turkey = (bacteria_total_mapped/turkey_total_mapped),
    unique_pct = 100-duplication
)
```

Add filter_status column

```{r macro_stats_add_filter_status, eval=FALSE}
# Load retained_samples_macro_filt_30 from counts data
load("data/macro/counts.Rdata")

# Add filter_status column to indicate which samples were retained by filtering
final_combined_stats_macro <- final_combined_stats_macro %>%
  mutate(filter_status = if_else(
    microsample %in% retained_samples_macro_filt_30,
    "Retained by filtering",
    "Excluded from filtering"
  ))
```

## Create plot_data_stats_macro

Combine sequencing statistics with sample metadata and alpha diversity for plotting.

Create plot_data_stats_macro with sequencing stats, metadata, and alpha diversity
```{r macro_stats_plot_data_stats, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# Load alpha diversity data (already loaded in load_data chunk)
# Combine sequencing stats with sample metadata and alpha diversity
plot_data_stats_macro <- final_combined_stats_macro %>%
  left_join(alpha_div_macro_filtered_30, by = join_by(microsample == microsample)) %>%
  left_join(sample_metadata_macro, by = join_by(microsample == sample))

```

## Save working objects

Save sequencing statistics data
```{r macro_stats_save, eval=FALSE}
# Save seq stats-related objects to separate file
save(
  final_combined_stats_macro,
  plot_data_stats_macro,
  file = "data/macro/seq_stats.Rdata"
)
```

## Validation

Validate sequencing statistics
```{r macro_stats_validation, eval=FALSE}
# Check dimensions
cat("Final combined stats dimensions:\n")
print(dim(final_combined_stats_macro))
cat("\nNumber of samples:", nrow(final_combined_stats_macro), "\n\n")

# Check that sanity checks pass
cat("Sanity check results:\n")
cat("Bacterial mapping methods agree:", all(comparison_bacteria_mapping_macro$same_counts), "\n")
if (!all(comparison_bacteria_mapping_macro$same_counts)) {
  cat("  Number of samples with differences:", sum(!comparison_bacteria_mapping_macro$same_counts), "\n")
  knitr::kable(comparison_bacteria_mapping_macro %>% 
    filter(!same_counts) %>%
    mutate(difference = bacteria_total_read_counts - bacteria_total_mapped) %>%
    select(microsample, bacteria_total_read_counts, bacteria_total_mapped, difference))
}
cat("\n")

# Verify percentages sum correctly (should be close to 100% for read composition)
cat("Percentage validation:\n")
# Check that percentages for read composition make sense
sample_check <- final_combined_stats_macro %>%
  mutate(
    total_pct_check = human_percentage + chicken_percentage + swine_percentage + 
                      turkey_percentage + histomonas_percentage + 
                      bacteria_percentage + unmapped_percentage
  ) %>%
  select(microsample, total_sequences_after_trim, total_pct_check, 
         human_percentage, chicken_percentage, swine_percentage, 
         turkey_percentage, histomonas_percentage, bacteria_percentage, unmapped_percentage)

cat("Mean total percentage (should be ~100):", mean(sample_check$total_pct_check, na.rm = TRUE), "\n")
cat("Range of total percentage:", range(sample_check$total_pct_check, na.rm = TRUE), "\n\n")

# Check for anomalies
cat("Anomaly checks:\n")
# Samples with very low bacterial percentage
low_bacteria <- final_combined_stats_macro %>%
  filter(bacteria_percentage < 1) %>%
  select(microsample, bacteria_percentage, total_sequences_after_trim)
if (nrow(low_bacteria) > 0) {
  cat("Samples with <1% bacterial reads:", nrow(low_bacteria), "\n")
  knitr::kable(head(low_bacteria, 10))
} else {
  cat("OK: No samples with <1% bacterial reads\n")
}
cat("\n")

# Samples with very high unmapped percentage
high_unmapped <- final_combined_stats_macro %>%
  filter(unmapped_percentage > 50) %>%
  select(microsample, unmapped_percentage, total_sequences_after_trim)
if (nrow(high_unmapped) > 0) {
  cat("Samples with >50% unmapped reads:", nrow(high_unmapped), "\n")
  knitr::kable(head(high_unmapped, 10))
} else {
  cat("OK: No samples with >50% unmapped reads\n")
}
cat("\n")

# Summary statistics
cat("Summary statistics:\n")
cat("  Mean bacterial percentage:", mean(final_combined_stats_macro$bacteria_percentage, na.rm = TRUE), "%\n")
cat("  Mean unmapped percentage:", mean(final_combined_stats_macro$unmapped_percentage, na.rm = TRUE), "%\n")
cat("  Mean histomonas percentage:", mean(final_combined_stats_macro$histomonas_percentage, na.rm = TRUE), "%\n")
cat("  Mean turkey percentage:", mean(final_combined_stats_macro$turkey_percentage, na.rm = TRUE), "%\n")
```

