# Beta Diversity - Macro Samples

```{r load_data_beta_macro, eval=FALSE}
load("data/macro/sample_metadata.Rdata")
load("data/macro/counts.Rdata")
load("data/micro/sample_metadata.Rdata")
load("data/MAG_catalogue/data.Rdata")
load("data/data_colors.Rdata")
```

## Prepare tidy tables for plotting

Create tidy tables for filtered genome counts (macro)
```{r tidy_plot_genome_counts_filt_30_beta_macro, warning=FALSE, comments="", message=FALSE, eval=FALSE}
tidy_plot_genome_counts_filt_30_macro <-  genome_counts_macro_filt_30 %>%
    pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
    left_join(genome_metadata, by = join_by(genome == genome)) %>% 
    left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>% 
    filter(count > 0) %>%
    mutate(phylum = factor(phylum, levels = phylum_level_vector),
           order = factor(order, levels = order_level_vector))

tidy_plot_genome_counts_filt_30_macro_closed <-  genome_counts_macro_filt_30 %>%
    mutate_at(vars(-genome), ~ . / sum(.)) %>% 
    pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
    left_join(genome_metadata, by = join_by(genome == genome)) %>% 
    left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>% 
    filter(count > 0) %>%
    mutate(phylum = factor(phylum, levels = phylum_level_vector),
           order = factor(order, levels = order_level_vector))
```



```{r plot_beta_macro_order_composition, warning=FALSE, comments="", message=FALSE, fig.height=7, fig.width=12}
tidy_plot_genome_counts_filt_30_macro %>%
  filter(treatment != "TM0") %>%
  ggplot(aes(x = animal , y = count, fill = order)) +
  geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
  scale_fill_manual(values = order_colors[-4], drop = FALSE) +
  labs(x = "Metagenomic Samples", y = "Order composition",fill = "Read type") +
  facet_nested(. ~ treatment_expl + age_category, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        axis.text.y = element_blank(),
        legend.position = "none") +
  ggtitle("Composition of bacterial orders at 30% coverage filtered data")

tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment != "TM0") %>%
  ggplot(aes(x = animal , y = count, fill = order)) +
  geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
  scale_fill_manual(values = order_colors[-4], drop = FALSE) +
  labs(x = "Metagenomic Samples", y = "Order composition",fill = "Read type") +
  facet_nested(. ~ treatment_expl + age_category, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        axis.text.y = element_blank(),
        legend.position = "none") +
  ggtitle("Composition of bacterial orders at 30% coverage filtered data")
```



## Species level
The problem here is that at species level there are too many zeros and then we need to drop almost all samples from Histomonas treatment after day a.
```{r plot_beta_macro_species_pca_prep, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10}
df_wide <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  # Remove metadata
  select(genome, microsample, count) %>%
  # Make tidy df into wide df
  pivot_wider(
    names_from = genome,   
    values_from = count,   
    # Fill missing values with 0
    values_fill = 0) %>%
  # Remove taxa (columns) that are zero for all leftover samples (rows)
  select(where(~ any(. != 0))) %>%
  # scale each sample to 100 (= closure to 100, not required, but common)
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100 )) %>%
  # turn the column 'microsample'into the name of the rows of the df
  column_to_rownames(var = "microsample")


# Remove taxa in few samples, and samples with few taxa
df_wide <- remove_samples_or_taxa(df = df_wide,
                                     min_samples_per_taxon = 15,
                                     min_taxa_per_sample = 20)


MACRO_sample_metadata <- sample_metadata_macro %>%
  mutate(microsample = sample) %>%
  select(microsample, animal, age_category, treatment, treatment_expl) %>%
  mutate(sample_type = "MACRO") 



pca_result <- perform_pca(df_wide, z_delete = TRUE)
p <- plot_pca(pca_result$pca_result,
         samples_color_metadata = "treatment" , 
         samples_shape_metadata = "age_category",
         samples_color_value = treatment_colours_bright, 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "species",
         MACRO_sample_metadata, genome_metadata, order_colors, 
         custom_ggplot_theme)

p2 = p + ggtitle("XX")
p2

```

```{r plot_beta_macro_species_pca_centroids, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10}

p <- plot_pca(pca_result$pca_result,
         samples_color_metadata = "treatment" , 
         samples_shape_metadata = "age_category",
         samples_color_value = treatment_colours_bright, 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "species",
         scaling_factor_value = 1.0, 
          loadings_number = 10,
         MACRO_sample_metadata, genome_metadata, order_colors, 
         custom_ggplot_theme,
         add_centroids = TRUE)

p2 = p + ggtitle("XX")
p2

```

## Genus level
Even here, the zero replacement needs to drop the samples with too few taxa. 
```{r plot_beta_macro_genus_pca_prep, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10}
genus_metadata <-  genome_metadata %>%
  select(domain, phylum, class, order, family, genus) %>%
  distinct(genus, .keep_all = TRUE) %>%
  mutate(genome = genus)

df_wide <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  # Remove metadata
  select(domain, phylum, class, order, family, genus, genome, microsample, count) %>%
  group_by(microsample, genus) %>%
  summarise(genus_count = sum(count, na.rm = TRUE))%>%
  filter(genus != "") %>%
  mutate(count=genus_count) %>%
  select(-genus_count) %>%
  pivot_wider(names_from = genus,values_from = count,values_fill = 0) %>%
  select(where(~ any(. != 0)) | where(is.character) | where(is.factor)) %>%   
  column_to_rownames(var = "microsample") %>%
  mutate(across(where(is.numeric), ~ . / rowSums(across(where(is.numeric))) * 100))


# Remove taxa in few samples, and samples with few taxa
df_wide <- remove_samples_or_taxa(df = df_wide,
                                     min_samples_per_taxon = 5,
                                     min_taxa_per_sample = 5)


MACRO_sample_metadata <- sample_metadata_macro %>%
  mutate(microsample = sample) %>%
  select(microsample, animal, age_category, treatment, treatment_expl) %>%
  mutate(sample_type = "MACRO") 



pca_result <- perform_pca(df_wide, z_delete = TRUE)
p <- plot_pca(pca_result$pca_result,
         samples_color_metadata = "treatment" , 
         samples_shape_metadata = "age_category",
         samples_color_value = treatment_colours_bright, 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "genus",
         sample_metadata = MACRO_sample_metadata, 
         genome_metadata = genus_metadata, 
         order_colors = order_colors, 
         custom_ggplot_theme)

p2 = p + ggtitle("XX")
p2

```
```{r plot_beta_macro_genus_pca_centroids, warning=FALSE, comments="", message=FALSE, fig.height=7, fig.width=9}

p <- plot_pca(pca_result$pca_result,
         samples_color_metadata = "treatment" , 
         samples_shape_metadata = "age_category",
         samples_color_value = treatment_colours_bright, 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "genus",
         sample_metadata = MACRO_sample_metadata, 
         genome_metadata = genus_metadata, 
         order_colors = order_colors, 
         scaling_factor_value = 1.5, 
          loadings_number = 10,
         custom_ggplot_theme,
         add_centroids = TRUE)

p2 = p + ggtitle("PCA at genus level")
p2

```


### PERMANOVA
```{r plot_beta_macro_permanova, warning = FALSE, comments = "", message = FALSE, fig.height = 7, fig.width=10}

# Calculate Euclidean distance on CLR data
pca_animals_dist_matrix <- vegdist(pca_result$df_clr_dist, method = "euclidean")

# Make sure that filtered df and metadata have the same number & order of rows
pca_animals_metadata_adonis <- plot_data_stats_macro %>%
  filter(microsample %in% rownames(pca_result$df_clr_dist)) %>%
  arrange(match(microsample, rownames(pca_result$df_clr_dist))) %>%
  select(microsample, animal, treatment, age_category)
all(pca_animals_metadata_adonis$microsample == rownames(pca_result$df_clr_dist))

# Run PERMANOVA
adonis2_result <- adonis2(
  pca_animals_dist_matrix ~ treatment + age_category,
  data = pca_animals_metadata_adonis,
  permutations = 999,
  by = "terms"
)

knitr::kable(adonis2_result)
```

```{r plot_beta_macro_permanova_by_treatment, warning = FALSE, comments = "", message = FALSE, fig.height = 7, fig.width=10}
# LOOP
treatments <- unique(pca_animals_metadata_adonis$treatment)

results_age <- map_dfr(treatments, function(trt) {
  
  # Subset metadata and restore proper sample IDs
  subset_meta <- pca_animals_metadata_adonis[
    pca_animals_metadata_adonis$treatment == trt, ,
    drop = FALSE
  ]
  rownames(subset_meta) <- subset_meta$microsample
  
  # Subset the distance matrix to those samples
  subset_dist <- as.dist(as.matrix(pca_animals_dist_matrix)[
    rownames(subset_meta),
    rownames(subset_meta)
  ])
  
  # Run PERMANOVA for age within this treatment
  adonis_res <- adonis2(subset_dist ~ age_category, data = subset_meta, permutations = 999)
  
  # Extract R² and p-value
  data.frame(
    Treatment = trt,
    R2 = adonis_res$R2[1],
    p_value = adonis_res$`Pr(>F)`[1]
  )
})

results_age

```



### Species representative genomes overlap
```{r plot_beta_macro_genomes_overlap_treatment, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=6}
group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", treatment == "TM1") %>%
  pull(genome) %>%
  unique()

group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", treatment == "TM2") %>%
  pull(genome) %>%
  unique()

group3 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", treatment == "TM3") %>%
  pull(genome) %>%
  unique()

group4 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", treatment == "TM4") %>%
  pull(genome) %>%
  unique()

venn_data <- list(
  TM1 = group1,
  TM2 = group2,
  TM3 = group3,
  TM4 = group4
)

# # Taxa unique to group1 only
# unique_group1 <- setdiff(group1, group2) # layer
# # Taxa unique to group2 only
# unique_group2 <- setdiff(group2, group1) # broiler
# 
# 
# # Filter once (layer/broiler + genomes of interest)
# filtered_group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
#   filter(type_simple == "P", animal_breed == "layer", genome %in% unique_group1)
# 
# filtered_group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
#   filter(type_simple == "P", animal_breed == "broiler", genome %in% unique_group2)
```

```{r plot_beta_macro_venn_treatment, message=FALSE, warning=FALSE, fig.height=6, fig.width=10}

ggVennDiagram(
  venn_data,
  label = "count",       
  set_color = c(TM1 = "#4059AE",
  TM2 = "#6A9AC3",
  TM3 = "#97D8C4",
  TM4 = "#F3B942") ) +
  scale_fill_gradient(low = "white", high = "grey") #+
  #ggtitle("Samples:") +
  #coord_flip()
```


```{r plot_beta_macro_genomes_overlap_age, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=6}
group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", age_category == "a") %>%
  pull(genome) %>%
  unique()

group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", age_category == "b") %>%
  pull(genome) %>%
  unique()

group3 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", age_category == "c") %>%
  pull(genome) %>%
  unique()

group4 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", age_category == "d") %>%
  pull(genome) %>%
  unique()

venn_data <- list(
  day_a = group1,
  day_b = group2,
  day_c = group3,
  day_d = group4
)

# # Taxa unique to group1 only
# unique_group1 <- setdiff(group1, group2) # layer
# # Taxa unique to group2 only
# unique_group2 <- setdiff(group2, group1) # broiler
# 
# 
# # Filter once (layer/broiler + genomes of interest)
# filtered_group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
#   filter(type_simple == "P", animal_breed == "layer", genome %in% unique_group1)
# 
# filtered_group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
#   filter(type_simple == "P", animal_breed == "broiler", genome %in% unique_group2)
```

```{r plot_beta_macro_venn_age, message=FALSE, warning=FALSE, fig.height=6, fig.width=10}

ggVennDiagram(
  venn_data,
  label = "count",       
  set_color = c(day_a = "#498a77",
  day_b = "#cc0100",
  day_c = "#cf4a82",
  day_d = "#b4a7d6") ) +
  scale_fill_gradient(low = "white", high = "grey") #+
  #ggtitle("Samples:") +
  #coord_flip()
```

